Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-12-20 ms.qmd
Fixed the figure
Added changes to the qmd

$R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures ([@fig-concept]A). For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$ and $lnVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes indicate that the mean (i.e., $lnRR_{Q_{10}}$) or variance ($lnVR_{Q_{10}}$) is larger at the higher of the two temperatures (numerator) when standardized to 10°C. Importantly, our effect sizes, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in variance) is log-linear (see [@fig-concept]B & *Supplementary Materials* for further discussion). We test and control for any violations of these assumptions in our analysis (see below).

@@

In addition to the acclimation period, all our models corrected for possible violations of the log-linearity assumption associated with effect size calculations ([@fig-concept]; and see Supplementary Materials @fig-lnrrq10_temp). We predicted that, if $lnRR_{Q_{10}}$ and $lnVR_{Q_{10}}$ were not strictly log-linear there would be a decrease in average effect size for studies applying higher temperature treatments, because these temperatures are expected to either converge on or cross the thermal maxima of the performance curve causing reaction rates to decelerate or decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of this approach is that we could still use $Q_{10}$ as an effect size while statistically correcting for the potential non-linearity that would be expected in the data at high treatment temperatures. Given that our data included a wide range of species and habitats, we also included a random slope of maximum temperature that varied across species because we expected that species would vary in their thermal performance curves, which would be reflected in experimental treatments. We mean-centered the maximum temperature and included it in our models.

With acclimation time and maximum temperature as moderators in all our models we proceeded to build separate models that tested our core questions. All estimates from our models are therefore conditioned on an average acclimation time (i.e., `r mean(data_wide$acclim_period, na.rm = TRUE)` days) and an average maximum temperature (i.e., 23°C) across the dataset.

@@

  scale_fill_manual(values = col) + scale_color_manual(values = rep("black",3)) + ylim(-1.2, 1.2)

@@

>>>>

$R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures ([@fig-concept]A). For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$ and $lnVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes indicate that the mean (i.e., $lnRR_{Q_{10}}$) or variance ($lnVR_{Q_{10}}$) is larger at the higher of the two temperatures (numerator) when standardized to 10°C. When measuring plasticity, it is the difference between $lnRR_{Q_{10}}$ acute (denoted, $lnRR_{Q_{10_{acute}}}$) and acclimation (denoted, $lnRR_{Q_{10_{acclimation}}}$) that captures the degree to which organisms plastically adjust (or acclimate). Importantly, our effect sizes, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in variance) is log-linear (see [@fig-concept]B & *Supplementary Materials* for further discussion). We test and control for any violations of these assumptions in our analysis (see below).

@@

In addition to the acclimation period, all our models corrected for possible violations of the log-linearity assumption associated with effect size calculations ([@fig-concept]; and see *Supplementary Materials* @fig-lnrrq10_temp). We predicted that, if $lnRR_{Q_{10}}$ and $lnVR_{Q_{10}}$ were not strictly log-linear there would be a decrease in average effect size for studies applying higher temperature treatments, because these temperatures are expected to either converge on or cross the thermal maxima of the performance curve causing reaction rates to decelerate or decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of this approach is that we could still use $Q_{10}$ as an effect size while statistically correcting for the potential non-linearity that would be expected in the data at high treatment temperatures. Given that our data included a wide range of species and habitats, we also included a random slope of maximum temperature that varied across species because we expected that species would vary in their thermal performance curves, which would be reflected in experimental treatments. We mean-centered the maximum temperature and included it in our models.

Lastly, all models included a random slope of effect type (acute vs acclimation) to estimate the variance in  the magnitude of plastic changes (acute vs acclimation) across studies. Such an analysis is similar to analyses using an effect size that is a contrast between $lnRR_{Q_{10_{acute}}}$ and $lnRR_{Q_{10_{acclimation}}}$ but is more powerful because it allows studies without acute responses to be included (see *Supplementary Materials*). With acclimation time and maximum temperature as moderators in all our models. Accounting for these in our meta-regression models, we proceeded to build separate models that tested our core questions. All estimates from our models are therefore conditioned on an average acclimation time (i.e., `r mean(data_wide$acclim_period, na.rm = TRUE)` days) and an average maximum temperature (i.e., 23°C) across the dataset.

@@

  scale_fill_manual(values = col) + scale_color_manual(values = rep("black",3)) + ylim(-1.2, 2)

@@

#### **Interaction-based effect sizes**

When measuring plasticity, what is reletavent is the difference between  lnRR_(Q_10 )acute and acclimation that captures the degree to which organisms plastically adjust (or acclimate).
